ExaModels and MadNLP: Open-Source Software Infrastructure for Accelerated Nonlinear Optimization on GPUs

### 2023-08-22 coin-or/main.tex
coin-or nomination added

We are delighted to submit our nominations for the COIN-OR Cup,
highlighting our contributions of {\bf ExaModels} and {\bf MadNLP} in the field of
computational infrastructure for operations research, specifically in
the area of nonlinear programming (NLP). Our open-source software
packages represent a significant breakthrough in efficiently solving
large-scale NLPs by leveraging the capabilities of modern GPU
hardware.  In this document, we address the primary challenges and
opportunities for optimization software development in 2023,
particularly within the context of accelerated computing. We elucidate
how ExaModels and MadNLP collectively address technical obstacles and
capitalize on the advantages offered by modern accelerated
computing. Furthermore, we emphasize the remarkable performance
enhancements achieved in solving real-world nonlinear optimization
problems at scale. A noteworthy example of such progress is our recent
accomplishment in accelerating alternating current (AC) optimal power
flow (OPF) problems by up to ten times compared to state-of-the-art
tools. Finally, we discuss how our development lays the foundation for
the advancement in various topics in operations research and unlocks
new possibilities.

\paragraph{Challenges and Opportunities}
In recent years, the scalability of deep learning has experienced
significant advancements, primarily due to the utilization of NVIDIA
GPUs, which have demonstrated remarkable success in constructing
large-scale AI models. Additionally, in the public domain, GPU
accelerators have provided a substantial portion of the computational
power for next-generation leadership HPC systems such as Frontier and
Aurora, achieving the milestone of one exaflop per second.  While GPUs
have proven highly effective in various scientific computing areas
like machine learning and high-fidelity simulations, integrating them
into constrained nonlinear programming (NLP) algorithms has encountered
considerable challenges. Specifically, the seamless translation of
sparse automatic differentiation (AD) and sparse linear solver
routines to GPU architecture has been recognized to be particularly
challenging.  This is because conventional sparse AD and matrix
factorization algorithms are designed for CPUs and heavily rely on
serial computation, making them unsuitable for the single-instruction,
multiple-data (SIMD) parallelism employed by GPU architectures (see
Figure \ref{fig:simd}, left).

>>>>

We are delighted to submit our nominations for the COIN-OR Cup,
highlighting our contributions of {\bf ExaModels} \cite{examodels} and
{\bf MadNLP} \cite{madnlp} in the field of computational
infrastructure for operations research, specifically in the area of
nonlinear programming (NLP). Our open-source software packages
represent a significant breakthrough in efficiently solving
large-scale NLPs by leveraging the capabilities of modern GPU
hardware.  In this document, we address the primary challenges and
opportunities for optimization software development in 2023,
particularly within the context of accelerated computing. We elucidate
how ExaModels and MadNLP collectively address technical obstacles and
capitalize on the advantages offered by modern accelerated
computing. Furthermore, we emphasize the remarkable performance
enhancements achieved in solving real-world nonlinear optimization
problems at scale. A noteworthy example of such progress is our recent
accomplishment in accelerating alternating current (AC) optimal power
flow (OPF) problems by up to ten times compared to state-of-the-art
tools. Finally, we discuss how our development lays the foundation for
the advancement in various topics in operations research and unlocks
new possibilities.

\paragraph{Challenges and Opportunities} In recent years, the
scalability of deep learning has experienced significant advancements,
primarily due to the utilization of NVIDIA
GPUs, which have demonstrated remarkable success in constructing
large-scale AI models. Additionally, in the public domain, GPU
accelerators have provided a substantial portion of the computational
power for next-generation leadership HPC systems such as Frontier and
Aurora, achieving the milestone of one exaflop per second.  While GPUs
have proven highly effective in various scientific computing areas
like machine learning and high-fidelity simulations, integrating them
into constrained NLP algorithms has encountered
considerable challenges. Specifically, the seamless translation of
sparse automatic differentiation (AD) and sparse linear solver
routines to GPU architecture has been recognized to be particularly
challenging.  This is because conventional sparse AD and matrix
factorization algorithms are designed for CPUs and heavily rely on
serial computation, making them unsuitable for the single-instruction,
multiple-data (SIMD) parallelism employed by GPU architectures (see
Figure \ref{fig:simd}, left).
